Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory

### 2022-12-05 Manuscripts/deNovoRates.tex
modified:   Modeling/DeNovoProb.jl

\title{Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory}
\author{Bharat Ravi Iyengar, Erich Bornberg-Bauer}
\date{\small Institute for Evolution and Biodiversity, Westfalian Wilhelms -- University of M\"{u}nster, H\"{u}fferstrasse 1, 48149 M\"{u}nster, Germany}

@@

\caption{Mutation bias and GC-content shape protein composition. Panels \textbf{(A)}, \textbf{(B)} and \textbf{(C)} show expected frequency, gain probability (\autoref{aagaineq}), and loss probability (\autoref{aalosseq}) for different amino acids (horizontal axes). The vertical axis in all these three subplots show the normalized values of these measures such that their sum over all amino acids is equal to one. Panel \textbf{(D)} shows the directionality (colored tiles, $\textit{D}_{x\to y}$, \autoref{directionality}) of the substitution from an amino acid (vertical axis) to another (horizontal axis). Positive values of $\textit{D}_{x\to y}$ are shown in shades of red, whereas negative values are shown in shades of blue. In all panels, amino acids are ordered based on their chemical similarity \citep{PMBEC}.}
\label{aafreq}

@@

We found that $\sim$37.5\% of amino acids in a protein sequence are expected to be hydrophobic. This finding is broadly in agreement with the previous study that places this percentage at 40\% \citep{hydrophobicRatchet}. Next, we asked if a protein sequence in general tends to be hydrophobic. Specifically, we calculated expected hydrophobicity ($\bar{\alpha}$, \autoref{expHydropathy}) of a protein sequence, based on expected frequency of different amino acids ($P_x$, \autoref{aafreq}) and their hydrophobicity values ($\alpha_x$).

@@

We show that genes can be lost much more rapidly than they are gained, such that a gene can be independently lost two to three times in the time required for it to emerge. Therefore, presence of a \textit{de novo} gene in one species (or population) cannot be interpreted as an evidence of gene emergence in this species.

Witin the limitations of our model assumptions, we answer the long standing question of which trajectory of mutational events leads to \textit{de novo} emergence. That is, whether ORF emerges first or transcription emerges first \citep{EBB-F1000}. We show that, in the absence of selection, long ORFs are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus. Long non-coding RNAs may be a potential source of \textit{de novo} genes with long ORFs but this evolutionary process may be constrained by a different set of factors. A previous study has shown that new transcripts frequently arise in regions overlapping existing genes. These new transcripts can harbor ORFs which in turn leads to \textit{de novo} gene emergence \citep{albaYeastdenovo}. This is an example of RNA-first trajectory. However, the ORFs present in these genes may be under indirect selection pressure that acts on the overlapping gene sequences. Therefore, they may be less frequently mutated or lost.

@@

We show that transcription loss is less likely than the loss of long ORFs. However, it may be difficult to find direct evidence for such a phenomenon. Specifically, it is difficult to infer from data if transcription was lost or never emerged. If the protein encoded by an ORF is indeed toxic, it would be silenced rapidly in microevolutionary time scales. One way of inferring negative selection from transcription loss may involve correlation of ORF sequence variation and transcription status. Mutations within ORF sequence are much more probable than mutations that disrupt transcription or ORF. This is so because the entire ORF sequence has higher number of mutable sites. If an ORF variant is untranscribed in many populations (species or taxa), it is possibly toxic.

Finally, we show that some amino acid substitutions are more likely to occur than others, and frequent substitutions that are expected to occur less frequently under neutral model, can suggest directional evolution.

Like all theoretical models, our models are not without limitations. First, we assume that nucleotides are uniformly distributed according to GC-content. This is not true as nucleotide composition can vary significantly across the genome. Mutation rates are also not uniform across the genome and can vary significantly \citep{mutbiasArabidopsis}. However, our models can be adapted to investigate specific genomic loci, where information on the local GC-content and mutation rate can more accurately determine the \textit{de novo} gene evolution in these loci. Second, our models are based on parameters such as mutation rate and mutation bias, which we obtained from published data on human and \textit{D.melanogaster} genomes \citep{humanmutrate,drosophilamutrate}. The accuracy of the models' prediction will depend on the accuracy of parameters. However, mutation rate itself only determines the values of different probabilities. Our results, which are based on probability ratios, will not significantly change with a different value of mutation rate. More importantly, different species may have a different spectrum of mutations \citep{joshmutbias}, especially for species that are highly distant from each other. Thus our inferences based on our chosen parameters will not universally apply to all species. Third, we assume that all mutations are independent. This is not the case for tandem mutations, where multinucleotide mutations occur more frequently than their expected frequency under the assumption of mutational independence \citep{MNM}. Finally, we ignore insertions, deletions and transpositions as mechanisms of mutation. Insertions and deletions (indels) are reported to be less frequent than substitutions \citep{drosophilamutrate}, and both indels and transpositions cause larger sequence alterations than point mutations. However, our modeling methodology allows incorporation of these mechanisms as a source of mutations.

A famous aphorism about theoretical models says ``all models are wrong but some are useful'' \citep{GEBox}. This is true for our models too. They are not 100\% accurate but they are useful for making predictions that can lead to a more focused experimental validation. These models can be made more complex to accommodate diverse molecular mechanisms driving gene evolution. Therefore our work opens up an opportunity for theoretical and computational biologists to develop models using our modeling framework, that more accurately describe their system of interest.

@@

Because a feature set usually has many DNA sequences, a mutation can change a feature sequence such that the resulting sequence is also a part of the feature set. Thus we defined the probability ($P_\textit{feature-stay}$) that a feature does not erode because of mutations. Specifically, it is the the sum of two probabilities. First is the probability that no mutation occurs in the DNA sequence ($P_0$), and the second probability describes the event where mutated sequence is still a part of the feature set.

>>>>

\title{Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory}
\author{Bharat Ravi Iyengar$^1$, Erich Bornberg-Bauer$^{1,2}$}
\date{\small $^1$Institute for Evolution and Biodiversity, Westfalian Wilhelms -- University of M\"{u}nster, H\"{u}fferstrasse 1, 48149 M\"{u}nster, Germany\\ $^2$Max Planck-Institute for Biology T\"{u}bingen, T\"{u}bingen, Germany}

@@

\caption{Mutation bias and GC-content shape protein composition. Panels \textbf{(A)}, \textbf{(B)} and \textbf{(C)} show expected frequency, gain probability (\autoref{aagaineq}), and loss probability (\autoref{aalosseq}) for different amino acids (horizontal axes). The vertical axis in all these three subplots show the normalized values of these measures such that their sum over all amino acids is equal to one. Panel \textbf{(D)} shows the directionality (colored tiles, $\textit{D}_{x\to y}$, \autoref{directionality}) of the substitution from an amino acid (vertical axis) to another (horizontal axis). Positive values of $\textit{D}_{x\to y}$ are shown in shades of red, whereas negative values are shown in shades of blue. We only show substitutions that can be realized by a single nucleotide mutation but we include all possible nucleotide mutations in the directionality calculations for these substitutions. In all panels, amino acids are ordered based on their chemical similarity \citep{PMBEC}.}
\label{aafreq}

@@

We found that $\sim$37.5\% of amino acids in a protein sequence are expected to be hydrophobic. This finding is broadly in agreement with the previous study that places this percentage at 40\% \citep{hydrophobicRatchet}. We note that cytosolic proteins may require more than 42\% of their constituent amino acids to be hydrophobic, in order to fold efficiently \citep{Dill1985}. Next, we asked if a protein sequence in general tends to be hydrophobic. Specifically, we calculated expected hydrophobicity ($\bar{\alpha}$, \autoref{expHydropathy}) of a protein sequence, based on expected frequency of different amino acids ($P_x$, \autoref{aafreq}) and their hydrophobicity values ($\alpha_x$).

@@

We show that genes can be lost much more rapidly than they are gained, such that a gene can be independently lost two to three times in the time required for it to emerge. In a hypothetical case, a proto-gene may be present in a species (A) but is absent in two sister taxa (B and C). Two evolutionary scenarios could explain this observation. In the first scenario, the proto-gene simply emerged for the first time in species A. The second scenario posits that the proto-gene was present in the common ancestor of the three species but was lost in both species B and species C. Our models suggest that the second scenario is more likely. Future phylogenetic studies on proto-genes should consider this finding for making inferences on the dynamics of gene gain and loss.

Within the limitations of our model assumptions, we answer the long standing question of which trajectory of mutational events leads to \textit{de novo} emergence. That is, whether ORF emerges first or transcription emerges first \citep{EBB-F1000}. We show that, in the absence of selection, long ORFs are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus. Long non-coding RNAs may be a potential source of \textit{de novo} genes with long ORFs but this evolutionary process may be constrained by a different set of factors. A previous study has shown that new transcripts frequently arise in regions overlapping existing genes. These new transcripts can harbor ORFs which in turn leads to \textit{de novo} gene emergence \citep{albaYeastdenovo}. This is an example of RNA-first trajectory. However, the ORFs present in these genes may be under indirect selection pressure that acts on the overlapping gene sequences. Therefore, they may be less frequently mutated or lost.

@@

We show that transcription loss is less likely than the loss of long ORFs. However, it may be difficult to find direct evidence for such a phenomenon. Specifically, it is difficult to infer from data if transcription was lost or never emerged. If the protein encoded by an ORF is indeed toxic, it would be silenced or purged rapidly in microevolutionary time scales. One way of inferring negative selection from transcription loss may involve correlation of ORF sequence variation and transcription status. Mutations within an ORF sequence are much more probable than mutations that disrupt transcription or an ORF. This is so because the entire ORF sequence can have a higher number of mutable sites than a promoter. If an ORF variant is untranscribed in many populations (species or taxa), it is possibly toxic.

A large volume of work exists on the likelihood of amino acid substitutions and many of these models are widely used for phylogenetics \citep{aasubOhta,PAM,blosum,submat92,submat92j,submat01,submat05,submat07,submat08}. However commonly used amino acid substitution models such as PAM and BLOSUM \citep{PAM,blosum} do not focus on directionality of mutations (the substitution matrices are symmetric). Our aim was to highlight how proteins can evolve under mutation pressure and that mutation bias causes some substitutions to occur more frequently than the others. Furthermore, we do not base our substitution probability calculations on known protein sequences but rather on random protein sequences that are more likely to to be encoded in proto-genes. Using our models, we show that for a pair of amino acids, one direction of substitution can be significantly more likely than the other. If a substitution occurs more frequently than it is expected to occur in our neutral model, then directional evolution can be a possible explanation for the observation. We also show that random mutations can lead to a small increase in a protein's hydrophobicity. Hydrophobic amino acids can facilitate folding \citep{Dill1985}, but can also cause proteins to aggregate \cite{hydrophobicRatchet}. Thus the effect of hydrophobicity increasing mutations on a protein's stability and toxicity, is dependent on where they occur in the protein structure.

Like all theoretical models, our models are not without limitations. First, we assume that nucleotides are uniformly distributed according to GC-content. This is not true as nucleotide composition can vary significantly across the genome. Mutation rates are also not uniform across the genome and can vary significantly \citep{mutbiasArabidopsis}. However, our models can be adapted to investigate specific genomic loci, where information on the local GC-content and mutation rate can more accurately determine the \textit{de novo} gene evolution in these loci. Second, our models are based on parameters such as mutation rate and mutation bias, which we obtained from published data on human and \textit{D.melanogaster} genomes \citep{humanmutrate,drosophilamutrate}. The accuracy of the models' prediction will depend on the accuracy of parameters. However, mutation rate itself only determines the values of different probabilities. Our results, which are based on probability ratios, will not significantly change with a different value of mutation rate. More importantly, different species may have a different spectrum of mutations \citep{joshmutbias}, especially for species that are highly distant from each other. Thus our inferences based on our chosen parameters will not universally apply to all species. Third, we assume that all mutations are independent. This is not the case for tandem mutations, where multinucleotide mutations occur more frequently than their expected frequency under the assumption of mutational independence \citep{MNM}. Fourth, we did not model the effect of many regulatory sequence features such as enhancers, transcription factor binding sites, 5' and 3' untranslated regions, and Kozak consensus sequence, that can influence the evolution of proto-genes. For example \textit{de novo} emergence is very likely to occur near enhancers \citep{PacoEnhancers}. We instead focused on minimal requirements for transcription and translation. However, our modeling framework allows inclusion of any number of sequence features in the calculations as long as their sequences can be defined (\autoref{methfeatures}). Finally, we ignore insertions, deletions and transpositions as mechanisms of mutation. Insertions and deletions (indels) are reported to be less frequent than substitutions \citep{drosophilamutrate}, and both indels and transpositions cause larger sequence alterations than point mutations. Our modeling methodology also allows incorporation of these mechanisms as a source of mutations.

A famous aphorism about theoretical models says ``all models are wrong but some are useful'' \citep{GEBox}. Our models are not an exception. They are not 100\% accurate but they are useful for making predictions that can lead to a more focused experimental validation. These models can be made more complex to accommodate diverse molecular mechanisms driving gene evolution. Therefore our work opens up an opportunity for theoretical and computational biologists to develop models using our modeling framework, that more accurately describe their system of interest.

@@

Because a feature set usually has many DNA sequences, a mutation can change a feature sequence such that the resulting sequence is also a part of the feature set. Thus we defined the probability ($P_\textit{feature-stay}$) that a feature does not erode because of mutations. Specifically, it is the sum of two probabilities. First is the probability that no mutation occurs in the DNA sequence ($P_0$), and the second probability describes the event where the mutated sequence remains a part of the feature set.
